---
title: "Linear Mixed Model for Exp 1"
author: "JF C"
date: "2025-11-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Linear Mixed Model for Accuracy (CR)

## Load R Packages

```{r}
library(lme4)
library(lmerTest)
library(emmeans)
library(tidyverse)
library(performance)
library(ggplot2)
library(lmtest)
```

## Load Data

```{r}
cr_wide <- read_csv("exp1_cr_wide.csv")
colnames(cr_wide)[1] <- 'sub'
```

## Data Transformation: From Wide to Long Format

We use pivot_longer to "collapse" all measurement columns.

names_to: Names of the new columns to be created.

names_sep: Separator used to split column names (e.g., "negative_invalid_1").
           Here we use "_" as the separator.

values_to: Name of the column storing measurement values (e.g., RT, Accuracy).

```{r ,echo=FALSE}
cr_long <- cr_wide %>%
  pivot_longer(
    cols = -c(sub),  # Select all columns except 'sub' (Subject ID) for transformation
    names_to = c("Emotion", "Validity", "Intensity"), # Split column names into three new columns
    names_sep = "_",                                   # Split by "_"
    values_to = "cr"                                   # New column name for the Dependent Variable (DV)
  )


# --- Data Cleaning (Important!) ---
cr_long <- cr_long %>%
  mutate(
    sub = factor(sub),
    Emotion = factor(Emotion),
    Validity = factor(Validity),
    Intensity = factor(Intensity)
  )

# Check the structure of the transformed long-format data
cat("\n--- Structure of Long-Format Data (str) ---\n")
str(cr_long)

```


## Model 0: Fixed Effects Only

```{r}
model_0 <- lm(
  cr ~ Emotion * Validity * Intensity,
  data = cr_long
)
print(anova(model_0))
```

## Model 1: Random Intercept Model

The DV is predicted by Emotion, Validity, Intensity, and all their interactions.

(1 | sub) indicates that we allow a random baseline intercept for each subject.

```{r}
cat("\n--- Fitting Model 1 (Random Intercept) ---\n")
model_intercept <- lmer(
  cr ~ Emotion * Validity * Intensity + (1 | sub),
  data = cr_long
)

cat("\n--- Summary of Model 1 ---\n")
print(summary(model_intercept))
```

## Model 2: Random Intercept and Random Slope Model

(1 + Intensity | sub) allows each subject to have a different intercept,

and allows the effect (slope) of Intensity to vary across subjects.

For demonstration, if Intensity is continuous (1, 2, 3, 4, 5), convert to numeric first:
`rt_long$Intensity_num <- as.numeric(as.character(rt_long$Intensity))`
`model_slope <- lmer(DV ~ Emotion * Validity * Intensity + (1 + Intensity_num | sub), data = rt_long)`

```{r}
cat("\n--- Fitting Model 2 (Random Slope) ---\n")
model_slope <- lmer(
  cr ~ Emotion * Validity * Intensity + (1 + Emotion+Validity+Intensity | sub),
  data = cr_long
)

cat("\n--- Summary of Model 2 ---\n")
print(summary(model_slope))
```

## Model Comparison
```{r comparison}
cat("\n--- Model Comparison (LRT) ---\n")
model_comparison0 <- lrtest(model_0, model_intercept, model_slope)
model_comparison <- anova(model_intercept, model_slope)
print(model_comparison0)
print(model_comparison)



# If Pr(>Chisq) is significant (e.g., < 0.05), then model_slope is better.
# Otherwise, stick to the more parsimonious model_intercept.
# We select one model uniformly for subsequent steps.
best_model <- model_slope 

# Estimate degrees of freedom using the Satterthwaite method to obtain F-values and p-values.
# This is similar to Type III Sum of Squares in traditional ANOVA.
cat("\n--- ANOVA Table for the Best Model (Type III F-tests) ---\n")
anova_table <- anova(best_model, type = "3", ddf = "Satterthwaite")
print(anova_table)



cat("\nComparison: Model 0 (No Random Effects), Model 1 (Random Intercept), Model 2 (Random Intercept & Slope)\n")
print(model_comparison0)
cat("\nComparison between Model 1 and Model 2\n")
print(model_comparison)
cat("\nANOVA Results for the Best Model\n")
print(anova_table)

```


# Linear Mixed Model for Reaction Time (RT)

## Import Data
```{r rt import data}
rt_wide <- read_csv("exp1_rt_wide.csv")
colnames(rt_wide)[1] <- 'sub'

rt_long <- rt_wide %>%
  pivot_longer(
    cols = -c(sub),  # Select all columns except 'sub' (Subject ID) for transformation
    names_to = c("Emotion", "Validity", "Intensity"), # Split column names into three new columns
    names_sep = "_",                                   # Split by "_"
    values_to = "rt"                                   # New column name for the Dependent Variable (DV)
  )


# --- Data Cleaning (Important!) ---
rt_long <- rt_long %>%
  mutate(
    sub = factor(sub),
    Emotion = factor(Emotion),
    Validity = factor(Validity),
    Intensity = factor(Intensity)
  )

# Check the structure of the transformed long-format data
cat("\n--- Structure of Long-Format Data (str) ---\n")
str(rt_long)


```

## Model 0
```{r}
rtmodel_0 <- lm(
  rt ~ Emotion * Validity * Intensity ,
  data = rt_long
)
print(anova(rtmodel_0))
```

## Model 1
```{r}
cat("\n--- Fitting Model 1 (Random Intercept) ---\n")
rtmodel_intercept <- lmer(
  rt ~ Emotion * Validity * Intensity + (1 | sub),
  data = rt_long
)

cat("\n--- Summary of Model 1 ---\n")
print(summary(rtmodel_intercept))
```

## Model 2
```{r}
cat("\n--- Fitting Model 2 (Random Slope) ---\n")
rtmodel_slope <- lmer(
  rt ~ Emotion * Validity * Intensity + (1 + Emotion+Validity+Intensity | sub),
  data = rt_long
)

cat("\n--- Summary of Model 2 ---\n")
print(summary(rtmodel_slope))
```

## Model Comparison
```{r}
cat("\n--- Model Comparison (LRT) ---\n")
model_comparison0 <- lrtest(rtmodel_0, rtmodel_intercept, rtmodel_slope)
model_comparison <- anova(rtmodel_intercept, rtmodel_slope)
print(model_comparison0)
print(model_comparison)



# If Pr(>Chisq) is significant (e.g., < 0.05), then model_slope is better.
# Otherwise, stick to the more parsimonious model_intercept.
# We select one model uniformly for subsequent steps.
best_model <- rtmodel_slope 

# Estimate degrees of freedom using the Satterthwaite method to obtain F-values and p-values.
# This is similar to Type III Sum of Squares in traditional ANOVA.
cat("\n--- ANOVA Table for the Best Model (Type III F-tests) ---\n")
anova_table <- anova(best_model, type = "3", ddf = "Satterthwaite")
print(anova_table)



cat("\nComparison: Model 0 (No Random Effects), Model 1 (Random Intercept), Model 2 (Random Intercept & Slope)\n")
print(model_comparison0)
cat("\nComparison between Model 1 and Model 2\n")
print(model_comparison)
cat("\nANOVA Results for the Best Model\n")
print(anova_table)
```

```{r check model, fig.width=12, fig.height=12}
cat("\n--- Model Diagnostics ---\n")
check_model(best_model)

```
